---
title: "Quantifying Seasonal Range Shifts with geoshift"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantifying Seasonal Range Shifts with geoshift}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set to TRUE once example data are bundled
)
#library(geoshift)
```

## Introduction

Seasonal migration is a consequential and widespread phenomenon in animal
ecology, yet characterizing the full scope of migratory diversity across
thousands of species remains a persistent challenge. Tracking technologies
provide the gold standard for documenting individual movement, but their
logistical and financial demands make it impractical to deploy them at the
taxonomic and geographic scales needed for broad comparative work. Incidental occurrence records 
from citizen science platforms and natural history collections have grown to the 
point where seasonally-explicit species distribution models (SDMs) are now feasible 
for an enormous range of taxa. This growth in data availability offers a path toward scalable, data-driven
inference of migratory behavior across the tree of life.

`geoshift` bridges the gap between SDM outputs and migration biology. Given
two seasonal SDM surfaces, the package provides a coordinated suite
of metrics that together characterize a species' migratory strategy. We
organize these metrics along three facets of migratory strategy:

- **Migration centroid distance**: the geographic distance (in km) between the
  centers of two seasonal distributions, used as a proxy for migration
  distance.
- **Seasonality**: the degree to which the species occupies distinct seasonal
  ranges versus a stable year-round range. This index ranges from −1 (entirely
  year-round, i.e., no seasonal shift) to +1 (seasonal distributions are
  entirely non-overlapping, i.e., a fully migratory species).
- **Range size change**: the ratio of range area in season 1 to season 2, capturing
  whether species contract into a smaller range, expand into a larger
  one, or maintain comparable range sizes across seasons.

Metrics to capture these facets of migratory strategy can be derived from both
continuous probability-of-occurrence surfaces and from binary
(presence/absence) surfaces, and `geoshift` provides dedicated functions for
each pathway. A set of complementary tools for working with raw occurrence
coordinates is also included for cases where point-based summaries are needed
alongside or instead of raster-based ones. Additionally, we provide a function
to generate centroid distance, seasonality, and range size change metrics from
two polygons representing seasonally-explicit distributions such as
published expert range maps.

---

## Installation and Setup

`geoshift` is currently available from GitHub:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("cjcampbell/geoshift")
```

Once installed, load the package alongside `terra` and `sf`, which provide the
spatial data classes that `geoshift` functions expect as inputs and return as
outputs:

```{r load}
library(geoshift)
library(terra)
library(sf)
```

---

## Input Data Requirements

`geoshift` is designed to work at the end of an SDM pipeline, after models have
been fit, projected, and, for polygon-based analyses, thresholded into a
binary presence/absence surface. The package supports four types of input data:

1. **Continuous surfaces** — raw probability-of-occurrence predictions output by
   most SDM algorithms (e.g., the cloglog output from Maxent). Values typically
   range from 0 to 1, representing relative habitat suitability or predicted
   occurrence probability.

2. **Binary (presence/absence) surfaces** — continuous surfaces that have been
   thresholded into predictions of presence (1) or absence (0), with cells not
   predicted to support occurrence set to NA or 0.

3. **Occurrence points** — occurrence records with x and y coordinate values
   denoting a recorded presence of a species.

4. **Polygons** — `sf` polygon objects representing a seasonally-explicit
   distribution, such as expert-drawn range maps.

Both raster surface types (continuous and binary) should share the same CRS,
extent, and resolution. The metrics that compare seasonal surfaces are only
interpretable if the two rasters are directly comparable in geographic space.

---

## Analyzing Continuous SDM Surfaces

Continuous SDM surfaces encode spatial variation in predicted occurrence
probability across every grid cell in the study extent. `geoshift` offers
complementary approaches to comparing a pair of such surfaces: a global overlap
statistic (Schoener's D), an ellipse-based geometric approach that characterizes
the shape, size, and position of each seasonal distribution, a tool for
examining SDM values at original occurrence locations, and a high-level
wrapper that bundles many of these statistics into a single call.

### Niche overlap with Schoener's D

Schoener's D is a widely used measure of niche overlap that quantifies the
degree to which two probability surfaces share the same spatial distribution of
predicted values. It is calculated as:

$$D = 1 - \frac{1}{2} \sum_{i} |p_{1,i} - p_{2,i}|$$

where $p_{1,i}$ and $p_{2,i}$ are the normalized predicted values for surface 1
and surface 2 in cell $i$. D ranges from 0 (no overlap — the surfaces predict
occurrence in completely different places) to 1 (perfect overlap — the surfaces
are identical in their spatial predictions). In the context of seasonal
migration, a low D value indicates that a species occupies very different
environments in its summer and winter seasons, while a high D suggests broadly
similar habitat use year-round.

`schoenersD()` takes two raster objects as inputs and returns a single numeric
value. Both surfaces are normalized internally before the calculation, so raw
values do not need to sum to 1 before being passed to the function.

```{r schoenersD}
d_value <- schoenersD(rast1 = summer_surface, rast2 = winter_surface)
d_value
```

### Visualizing where seasons differ: the Schoener's projection surface

A single D value summarizes the global degree of overlap between two surfaces,
but it cannot tell you *where* the surfaces agree or diverge geographically.
`schoenersProjection()` computes a cell-by-cell difference raster that
preserves this spatial information. Each cell in the output raster reflects how
much more strongly that cell is predicted under one surface versus the other,
after both have been normalized. Cells where the summer surface predominates
will have positive values; cells where the winter surface predominates will have
negative values; and cells where the two surfaces make similar predictions will
be near zero.

By default, `schoenersProjection()` returns absolute differences (`abs = TRUE`),
which is useful for identifying regions of high disagreement regardless of
direction. Setting `abs = FALSE` returns signed differences, which reveal
directionality — identifying which cells are distinctly "summer-season cells"
versus "winter-season cells."

```{r schoenersProj}
# Signed projection: positive = summer, negative = winter
similarity_rast <- schoenersProjection(rast1 = summer_surface,
                                       rast2 = winter_surface,
                                       abs   = FALSE)
```

This output raster can be passed directly to `makeCompoundPlot()` for
visualization (see Section 7), or explored with any standard raster plotting
workflow.

### Data ellipses: geometric summaries of seasonal distributions

While Schoener's D and the projection surface capture cell-level similarity,
ecologists often want a more geometric summary, the location, size, and
orientation of each seasonal distribution as a whole. `geoshift` implements a
weighted resampling ellipse approach for this purpose.

The core idea is straightforward: treat the continuous SDM surface as a sampling
weight and draw a large number of random points from the geographic domain of
the raster in proportion to predicted occurrence probability. The spatial
distribution of those resampled points is then summarized by a data ellipse
enclosing a specified proportion of points (controlled by `prop_points`, with a
default of 0.50). Because higher-probability cells contribute more points to
the sample, the ellipse reflects the "center of mass" of the predicted
distribution rather than its full spatial extent. This makes ellipses more
robust to spatially diffuse low-probability predictions at the fringes of the
modeled range than methods based on convex hulls or raw raster extents.

`surface2df()` is a utility function that converts a raster surface into a
three-column data frame of x coordinates, y coordinates, and cell values,
with NA cells removed. It is used internally by several `geoshift` functions
and can also be called directly when a data frame representation of a raster is
needed.

`makeDataEllipse()` generates a single ellipse for one surface. It requires a
data frame of coordinates and their associated weights (cell values), a CRS
string, and the `prop_points` parameter:

```{r ellipse-single}
# Convert raster to a weighted coordinate data frame first
summer_df <- surface2df(summer_surface)

summer_ellipse <- makeDataEllipse(
  coords      = summer_df[, c("x", "y")],
  weights     = summer_df$value,
  crsString   = crs(summer_surface),
  prop_points = 0.50
)
```

Each ellipse object is a list with two elements: `$sf` (an `sf` polygon, the
recommended format for downstream analysis) and `$st` (a `SpatialPolygons`
object, retained for compatibility).

For most workflows, `makeEllipses()` is more convenient. It is a wrapper that
calls `makeDataEllipse()` on both rasters at once and returns a named list of
two ellipses:

```{r ellipses-both}
my_ellipses <- makeEllipses(
  rast1       = summer_surface,
  rast2       = winter_surface,
  prop_points = 0.50,
  n           = 10000
)
```

The `n` argument controls the number of points resampled from each surface.
Larger values produce more stable ellipse estimates but increase computation
time; 10,000 is generally sufficient for smooth ellipses on continental-scale
rasters.

### Change at points: reading SDM values at occurrence locations

`calculateChangeAtPoints()` provides a complementary perspective on seasonal
distributional change by asking: for each occurrence point recorded in one
season, how suitable does the SDM predict that location to be in the *other*
season? This requires both raster surfaces and both sets of occurrence
coordinates, and returns a data frame with the raw SDM values from each surface
extracted at each point, along with a `difference` column encoding the
cross-season change.

For an occurrence point recorded in the first season, the difference value is
the second-season surface value minus the first-season surface value at that
location. A negative difference indicates that the first-season site is
predicted to become less suitable in the following season, consistent with a
species that departs that location. For a second-season occurrence point, the
sign is reversed: positive differences indicate that the second-season site is
predicted to have been less suitable during the first season — for example, a
winter-specialist site in a species with distinct seasonal ranges.

```{r changeAtPoints}
change_df <- calculateChangeAtPoints(
  rast1     = summer_surface,
  rast2     = winter_surface,
  rastnames = c("Summer", "Winter"),
  pts1      = summer_coords,
  pts2      = winter_coords
)
```

The output data frame retains the original coordinates of each point alongside
the extracted and difference values, making it straightforward to map the
spatial distribution of seasonal change across the observation record or to
summarize it with descriptive statistics (e.g., mean difference, proportion of
points where the winter surface value exceeds the summer surface value).

### Extracting the full statistics table from continuous surfaces

`extractStatistics()` is the high-level function for deriving a comprehensive
set of comparison metrics from two continuous raster surfaces in a single call.
Internally, it computes Schoener's D, constructs data ellipses for both
surfaces (or accepts pre-computed ellipses via `myEllipses`), and derives
centroid positions, centroid distance, bearing between centroids, ellipse areas,
and the proportion of ellipse overlap. The result is returned as a single-row
data frame that is easy to bind with results from other species or time periods.

The `rastnames` argument accepts any seasonal labels that are meaningful for
your study system — `c("Summer", "Winter")`, `c("Wet season", "Dry season")`,
`c("Breeding", "Nonbreeding")`, and so on.

```{r extractStats}
stats_table <- extractStatistics(
  rast1      = summer_surface,
  rast2      = winter_surface,
  rastnames  = c("Summer", "Winter"),
  species    = "Tadarida_brasiliensis",
  myEllipses = my_ellipses
)
```

Key columns in the output include:

- `schoenersD`: global niche overlap (0–1)
- `centroid.distance`: great-circle distance in km between the centroids of
  the two ellipses — the continuous-surface analog of migration distance
- `centroid.bearing`: compass bearing from the season-one centroid to the
  season-two centroid, useful for describing the direction of seasonal movement
- `Ellipse.Area`: area of each ellipse in km²
- `Area.Ratio`: ratio of ellipse areas (season one : season two)
- `percent.Ellipse.intersected`: the proportion of each ellipse overlapped by
  the other, a geometric analog of the seasonality index

If you are processing multiple species, pre-computing ellipses with
`makeEllipses()` and passing them via `myEllipses` skips the resampling step
and substantially reduces run time for batch workflows.

---

## Analyzing Binary (Presence/Absence) Surfaces

By collapsing the continuous probability surface into discrete predictions of
where a species is and is not expected to occur, binary surfaces yield range
polygons that can be compared geometrically — how much do they overlap? Which
parts are unique to each season? How far apart are the range centers?

### Importance of thresholding decisions

The translation from continuous probability to a discrete presence/absence
prediction involves a choice of threshold, and that choice is consequential.
The choice of thresholding method is upstream of `geoshift` and not performed
by the package, but it matters for the metrics you will derive. Automated
thresholding methods based on model performance statistics (e.g., maximizing
the True Skill Statistic, TSS) provide a reproducible, assumption-light
approach that works well when occurrence data are dense and model validation is
robust. Expert thresholding — in which the researcher inspects maps and adjusts
the threshold against independent knowledge of the species' range — can improve
accuracy when data are sparse or when automated methods produce implausible
predictions. In our validation study (Campbell et al., forthcoming),
expert-thresholded binary surfaces outperformed both automated binary metrics
and continuous-surface metrics for explaining benchmark measures of centroid
distance and seasonality across 50 North American bird species. We recommend
retaining and documenting the threshold value applied to each surface.

### The three core binary or presence/absence (PA) metrics

`extractPAstatistics()` is the primary function for the binary surface
workflow. It accepts two binary raster objects, converts them to `sf` polygon
objects, and passes them to `statsFromPolygons()` to compute the three core
metrics. The result is a data frame with column names prefixed by `"PA_"` by
default, containing:

**Centroid distance** (`PA_centroidDist_km`): the distance in kilometers
between the geographic centroids of the two range polygons. This is the most
direct SDM-based proxy for migration distance. A species with non-overlapping
summer and winter ranges on opposite sides of a continent will have a large
centroid distance; a year-round resident will have a centroid distance near
zero.

**Seasonality index** (`PA_yrRoundVseasonal`): a normalized index that
quantifies the degree to which the range is seasonal versus year-round. The
index is computed by comparing the area of the geometric intersection of the
two seasonal polygons (the year-round component of the range) against the areas
of the portions unique to each season. The index ranges from −1, indicating
that the entire predicted range is year-round (no seasonal shift), to +1,
indicating that the summer and winter ranges are entirely non-overlapping (a
fully migratory species with distinct seasonal destinations). Values near zero
indicate partial migration, in which part of the population or range is stable
year-round and part is seasonally distinct.

**Range size change** (`PA_rangeSizeChange`): the ratio of the area of the
first surface's polygon to that of the second (e.g., summer:winter area).
Values greater than 1 indicate a larger first-season range; values less than
1 indicate the opposite. This metric captures phenomena such as summer-season
range expansion (common in Temperate Zone breeders that winter in the tropics),
winter aggregation at productive wintering sites, or post-breeding range
contraction.

```{r extractPA}
pa_metrics <- extractPAstatistics(
  PA1 = summer_binary,
  PA2 = winter_binary
)
pa_metrics
```

### Working directly with polygon objects

If you already have `sf` polygon objects — for example, published range maps
from BirdLife International, NatureServe, or the IUCN Red List — you can pass
them directly to `statsFromPolygons()`. This is the underlying function called
by `extractPAstatistics()`, and it accepts any pair of `sfc_POLYGON` or
`sfc_MULTIPOLYGON` objects:

```{r statsFromPolygons}
polygon_metrics <- statsFromPolygons(
  sf1 = summer_polygon,
  sf2 = winter_polygon
)
```

The output column names from `statsFromPolygons()` do not carry the `"PA_"`
prefix. This makes it easy to compare metrics derived from SDM surfaces with
those from published range maps in the same data frame: the `"PA_"` prefix
keeps SDM-derived metrics distinct from polygon-derived metrics when columns
are joined.

### Computing range area

`areaOfExpectedOccurrence()` calculates the geographic area represented by
the presence cells in a binary raster. It reclassifies the surface to retain
only presence cells, converts the result to an `sf` polygon, and returns the
total area as a `units` object in the native units of the raster's CRS (most
commonly km² for projected rasters). This function is called internally by
`extractPAstatistics()` and `statsFromPolygons()`, but it is also available
directly for cases where standalone area calculations are needed — for example,
to report absolute range sizes alongside relative metrics.

```{r areaOfExpectedOccurrence}
summer_area <- areaOfExpectedOccurrence(summer_binary)
winter_area <- areaOfExpectedOccurrence(winter_binary)

# Area is returned as a units object; convert to numeric km² if needed
as.numeric(summer_area)
```

---

## Using Occurrence Point Data

`geoshift` provides two functions that convert occurrence coordinate data
frames into polygon objects. These polygons can then be passed to
`statsFromPolygons()` to obtain the same three core metrics (centroid distance,
seasonality, range size change) that are available from binary raster surfaces.
We note that our validation study found metrics derived from occurrence points
alone performed worse at describing migration strategies than those derived from
SDMs; point-based summaries are therefore most appropriate as exploratory
baselines or in cases where SDM fitting is not feasible.

### Convex hull polygons

A convex hull is the smallest convex polygon that encloses a set of points.
It is the simplest possible polygon summary of an occurrence dataset, requiring
no modeling assumptions. `coordsToConvexHull()` accepts a two-column data
frame, matrix, or `SpatialPoints` object of longitude–latitude coordinates and
an optional `proj4string` specifying the target CRS, and returns an `sf`
polygon. The returned polygon can be passed directly to `statsFromPolygons()`
for metric extraction.

```{r convexHull}
summer_hull <- coordsToConvexHull(
  coords      = summer_coords,
  proj4string = "+proj=longlat +datum=WGS84"
)

winter_hull <- coordsToConvexHull(
  coords      = winter_coords,
  proj4string = "+proj=longlat +datum=WGS84"
)

hull_metrics <- statsFromPolygons(sf1 = summer_hull, sf2 = winter_hull)
```

Convex hull polygons are sensitive to spatial outliers in occurrence records
and will tend to overestimate range size in datasets where records are unevenly
distributed across the true range. They are most appropriate as rapid
exploratory summaries or as baselines against which model-based estimates can
be compared.

### Kernel density estimate polygons

Kernel density estimation (KDE) offers a more nuanced coordinate-to-polygon
approach that is less sensitive to outliers and better captures the spatial
density of occurrence records. `coordsToKDEPolygon()` fits a bivariate kernel
density surface to the input coordinates and returns an `sf` polygon
representing the contour at a specified density percentile (the `percentile`
argument, defaulting to 50). A 50th-percentile polygon encloses the geographic
area where the modeled density of occurrence records is at or above the median —
roughly capturing the core of the observed range rather than its full extent.

```{r kdePolygon}
summer_kde <- coordsToKDEPolygon(
  coords      = summer_coords,
  percentile  = 50,
  proj4string = "+proj=longlat +datum=WGS84"
)

winter_kde <- coordsToKDEPolygon(
  coords      = winter_coords,
  percentile  = 50,
  proj4string = "+proj=longlat +datum=WGS84"
)

kde_metrics <- statsFromPolygons(sf1 = summer_kde, sf2 = winter_kde)
```

The `extent` argument controls the spatial extent of the grid used to compute
the kernel density surface and can be increased if the default grid clips the
density surface at its edges.

---

## Visualization

`geoshift` includes two purpose-built plotting functions that produce
`ggplot2`-based visualizations of seasonal SDM outputs. Both functions return
`ggplot` objects that can be further modified with standard `ggplot2` syntax
after they are created.

### Single-season maps with `makePlot()`

`makePlot()` produces a map of a single seasonal surface colored by a gradient
from light grey (low predicted occurrence) to a season-specific accent color
(high predicted occurrence). By convention, summer surfaces are rendered in red
and winter surfaces in blue, though the `maxCol` argument allows the accent
color to be overridden. The function accepts the three-column data frame output
of `surface2df()` rather than a raster directly, making the conversion step
explicit:

```{r makePlot}
summer_df <- surface2df(summer_surface)
winter_df <- surface2df(winter_surface)

p_summer <- makePlot(surface = summer_df,
                     species = "American Redstart",
                     season  = "Summer")

p_winter <- makePlot(surface = winter_df,
                     species = "American Redstart",
                     season  = "Winter")
```

The `coordArgs` parameter accepts a named list of arguments passed to
`coord_sf()`, which is useful for adjusting the geographic extent of the plot —
for example, cropping to a region of interest rather than the full raster
extent. `pngSaveFileName` optionally saves the plot directly to a PNG file.

### Compound comparison plots with `makeCompoundPlot()`

`makeCompoundPlot()` assembles a multi-panel visualization that places the full
seasonal comparison in a single figure. The central panel displays the
Schoener's projection surface — the signed cell-by-cell difference between the
two normalized seasonal surfaces — as a diverging color gradient from blue
(winter-dominant cells) through grey (similar predictions in both seasons) to
red (summer-dominant cells). Overlaid on this background are the two data
ellipses (red for `rast1`, blue for `rast2`) and their respective centroids,
providing an immediate visual summary of the shape and separation of the two
seasonal distributions.

An optional `summaryTable` argument accepts a data frame of metrics (e.g.,
the output of `extractStatistics()` or `extractPAstatistics()`) and renders it
as a formatted table below the map panel, making `makeCompoundPlot()` suitable
for generating a self-contained summary figure for each species in a
multi-species analysis.

```{r makeCompoundPlot}
compound_plot <- makeCompoundPlot(
  rast1        = summer_surface,
  rast2        = winter_surface,
  rastnames    = c("Summer", "Winter"),
  myEllipses   = my_ellipses,
  species      = "American Redstart",
  summaryTable = stats_table
)

compound_plot
```

As with `makePlot()`, `pngSaveFileName` can be used to write the result
directly to disk. Passing pre-computed ellipses via `myEllipses` avoids
re-running the resampling step if ellipses were already computed for
`extractStatistics()`.

---

## Interpreting Results in a Migration Biology Context

Migration centroid distance, seasonality index, and range size change are most
informative when interpreted together rather than in isolation.

A large centroid distance paired with a high seasonality index (near +1)
describes a strongly migratory species with geographically distinct seasonal
ranges — the pattern typical of long-distance Neotropical migrants in North
America that breed across temperate latitudes in summer and concentrate in
Central and South America in winter.

A small centroid distance paired with a seasonality index near zero describes a
resident or short-distance mover whose range center is stable across seasons.
Intermediate values of both metrics suggest a partial migrant: a species in
which some populations or individuals migrate while others remain on or near the
summer range.

Range size change adds further nuance. A summer:winter area ratio substantially
greater than 1 (summer range much larger) may indicate that a species
aggregates into a smaller, resource-rich winter range — a pattern common in
species that exploit seasonally productive temperate breeding habitats and then
concentrate on tropical or subtropical wintering grounds. A ratio near 1
suggests comparable seasonal range sizes, while a ratio less than 1 (summer
range smaller than winter range) may reflect a species that expands its range in
winter, perhaps tracking spatially distributed food resources across a broad
landscape.

A few additional considerations are worth keeping in mind when interpreting
`geoshift` outputs:

**Data density effects.** Metrics derived from SDMs built with few occurrence
records will have greater uncertainty than those based on dense occurrence
datasets. In the validation study accompanying this package, metrics derived
from models built with as few as 50 occurrence records per season remained
informative for centroid distance and seasonality, but range size change was
more sensitive to data density and should be interpreted more cautiously when
records are sparse.

**Thresholding sensitivity.** The choice of threshold for binary surfaces
directly affects the area and position of the resulting range polygons, and
therefore the polygon-based metrics. Comparing metrics across species
thresholded with different methods (e.g., automated TSS thresholding for
data-rich species and expert thresholding for data-poor ones) introduces a
potential source of systematic bias and should be avoided where possible.

**Spatial extent.** Predictions are only made within the geographic extent used
to project the SDM. If the study extent does not encompass a species' full
seasonal range — for example, if a species' winter range extends beyond the
study region — the centroid and polygon metrics will reflect only the modeled
portion of the range and may underestimate the true seasonal shift.

---

## Summary and Further Resources

`geoshift` provides a structured, reproducible workflow for translating seasonal
SDM outputs into ecologically interpretable metrics of migratory behavior,
making it possible to characterize migratory strategy across large numbers of
species in a consistent, comparable framework.

**Companion manuscript:** The methods implemented in `geoshift` are described
and validated against eBird abundance benchmarks across 50 North American bird
species in Campbell et al. (forthcoming). A link to the published paper and
associated preprint will be provided here upon availability.

**SDM fitting:** Methods describing the SDM pipeline used to generate the SDMs
in our companion manuscript are described in Abbott, J. C., Bota-Sierra, C. A.,
Guralnick, R., Kalkman, V., González-Soriano, E., Novelo-Gutiérrez, R., ... &
Belitz, M. W. (2022). Diversity of Nearctic dragonflies and damselflies
(Odonata). *Diversity*, 14(7), 575.

**Package source and issue tracker:**
<https://github.com/cjcampbell/geoshift>
